---
title: "Negative Controls for Differential Gene Expression Analysis and Gene Set Enrichment Analysis"
author: "Kieran Redpath"
format: html
editor: source
theme: cosmo
---

## Summary
+ Carries out differential expression analysis between BRCA and STAD cell lines with the top and bottom tertiles for DOI response.
+ Saves these outputs as .rds files for later use.
+ Reloads processed data from 0_Setup.qmd and 1_Differential_Expression_Analysis.qmd
+ Carries out gene set enrichment analysis for Reactome pathways on differentially expressed gene list, ranked by limma-moderated t-statistic.
+ Collapses Reactome pathway terms to shared parent pathway terms.
+ Saves these outputs as .rds files for later use.

### Load packages
```{r}
#| results: hide
#| warning: false
#| message: false
#| echo: false

library(ggplot2)
library(limma)
library(edgeR)
library(data.table)
library(magrittr)
library(CePa)
library(tidyverse)
library(ggbeeswarm)
library(dplyr)
library(ReactomePA)
library(org.Hs.eg.db)
library(reactome.db)
library(goseq)
library(gplots)
source('data/raw/heatmap-mik.R')
library(knitr)
library(kableExtra)
library(devtools)
library(viridis)
library(openxlsx)
library(fgsea)
set.seed(42)
```

### Load pre-processed data
```{r}
# GDSC DOI response data for cell lines used in DE analysis.
sort_DOI_GDSC <- readRDS("data/sort_DOI_GDSC.rds")

# CCLE normalized gene expression data for cell lines used in DE analysis.
sort_voom_CCLE <- readRDS("data/sort_voom_CCLE.rds")
```

### Examine the data
```{r}
# Check that all column names are identical between the CCLE and GDSC datasets
sum(colnames(sort_voom_CCLE)==sort_DOI_GDSC$CCLE_Name)
# What types of data are included in the processed GDSC dataset?
names(sort_DOI_GDSC)
```

### Randomise gene names for negative controls
```{r}
# Randomise gene names to run a negative control for pathway analysis.
sort_voom_CCLE_negative <- sort_voom_CCLE
# Create a randomised list of the gene names.
random_genes <- sample(rownames(sort_voom_CCLE_negative))
# Reassign these to the matrix.
rownames(sort_voom_CCLE_negative) <- random_genes

# Tidy up the environment
rm(list = c("random_genes"))
```

### Differential expression analysis
```{r}
# Create groups for comparison, based on drug response to DOI.
group <- cut(
  sort_DOI_GDSC$AUC,
  breaks = quantile(sort_DOI_GDSC$AUC, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
  labels = c("Bottom", "Middle", "Top"),
  include.lowest = TRUE
)

# How many cell lines are in each group?
table(group)
# Visualize the ranges of drug response values in each group.
ggplot(sort_DOI_GDSC, aes(x = group, y = AUC)) +
  geom_boxplot() +
  labs(title = "AUC Distribution by Group",
       x = "Group", y = "AUC") +
  theme_minimal()
# Create the design matrix.
design <- model.matrix(~group); colnames(design) <- c("Mean", "BottomVsMiddle", "BottomVsTop")
# Fit the linear model.
fit <- lmFit(sort_voom_CCLE_negative, design) %>% eBayes(.)
# Create the top table of DE genes, and ensure that it's sorted by adjusted p-value.
top_table <- topTable(fit, coef = "BottomVsTop", adjust = "BH", n = nrow(sort_voom_CCLE_negative)) %>%
  arrange(adj.P.Val)

# Tidy up the environment.
rm(list = c("group", "design", "fit"))
```

### Explore the differential expression data
```{r}
# Plot the counts distribution of the gene with the smallest adjusted p-value across these cell lines.
sort_voom_CCLE_negative[,1] %>% density(.) %>% plot(.)
# How many genes are significantly differentially expressed at BH-adjusted α = 0.05?
sum(top_table$adj.P.Val < 0.05)
# Which genes are significantly differentially expressed at BH-adjusted α = 0.05?
top_table %>% filter(adj.P.Val < 0.05) %>% .[1:10, ]

# Plot fold change in AUC against adjusted p-value.
# Create new columns in top_table for plotting.
top_table$logAdjP <- -log10(top_table$adj.P.Val)
top_table$highlight <- (top_table$adj.P.Val < 0.05)
# Define the adjusted p-value threshold line.
adj_p_thresh <- -log10(0.05)
# Plot differential expression for diagnosis.
volcano <- ggplot(top_table, aes(x = logFC, y = logAdjP)) +
  geom_point(aes(color = highlight), size = 1) +
  scale_color_manual(values = c("black", "red")) +
  geom_hline(yintercept = adj_p_thresh, linetype = "dashed", color = "blue") +
  labs(title = "Volcano Plot (Adjusted P-values)", x = "log2 Fold Change", y = "-log10(Adjusted P-value)") +
  theme_minimal()
# Print.
volcano

# Remove plotting columns from top_table.
top_table <- top_table %>% dplyr::select(-logAdjP, -highlight)

# Tidy up the environment.
rm(list = c("adj_p_thresh"))
```

### Load pre-processed data
```{r}
# Tool for converting gene names without calling AnnotationDbi::select() every time.
gene_tool <- readRDS("data/gene_tool.rds")

# Tool for indexing cell lines to tissues.
tissue_tool <- readRDS("data/tissue_tool.rds")
```

### Rank genes by limma-moderated t-statistic
```{r}
# Format genes for Reactome pathway analysis, converting to Entrez IDs for fgsea, while handling missing and duplicate Entrez IDs.
ranked_genes <- top_table %>%
  rownames_to_column(var = "ENSEMBL") %>% # Convert rownames to a column.
  arrange(adj.P.Val) %>% # Ensure that the top_table is sorted by BH-adjusted p-value.
  left_join(gene_tool, by = c("ENSEMBL" = "ENSEMBL")) %>% # Add information for Entrez ID and gene symbol.
  dplyr::filter(!is.na(ENTREZID)) %>% # Remove genes without Entrez gene IDs.
  distinct(ENTREZID, .keep_all = TRUE) %>% # In case of duplicates, keep only the first occurrence of each Entrez ID (i.e. the one with the lowest BH-adjusted p-value).
  dplyr::select(ENTREZID, t) %>% # Keep only Entrez IDs and t-statistic.
  arrange(t) %>% # Arrange by t-statistic for GSEA.
  deframe() # Convert to named vector.
```

### Carry out gene set enrichment analysis
```{r}
#| warning: false
# Suppress warning about ties in some gene ranks for GSEA.

# Find Reactome pathways that these genes are present in.
all_pathways <- ranked_genes %>% names(.) %>% reactomePathways(.)
# 1:many mapping reflects that many genes map to multiple Reactome pathways.

# Carry out GSEA analysis, excluding uninformative pathways with over 500 genes.
gsea_results <- fgsea(pathways = all_pathways, stats = ranked_genes, maxSize = 500) %>% 
  arrange(padj)

# Identify pathways that are significantly enriched at a given BH-adjusted α.
enriched_pathways <- gsea_results %>% dplyr::filter(padj < 0.05) %>% 
  arrange(padj)

# Identify shared parent pathways among significantly enriched pathways.
parent_pathway_mapping <- collapsePathways(enriched_pathways, all_pathways, ranked_genes)
parent_enriched_pathways <- enriched_pathways %>%
  dplyr::filter(pathway %in% parent_pathway_mapping$mainPathways) %>%
  arrange(padj)

# Tidy up the environment.
rm(list = c("ranked_genes", "all_pathways", "gsea_results", "parent_pathway_mapping"))
```

### Inspect significantly enriched Reactome pathways
```{r}
# Reactome pathways
enriched_pathways
# Parent Reactome pathways
parent_enriched_pathways

# Print a summary message
paste0(
  sum(top_table$adj.P.Val < 0.05), 
  " genes are significantly differentially expressed with randomised IDs. ",
  "After GSEA, ", 
  nrow(enriched_pathways), 
  " Reactome pathways were significantly enriched in this gene list. ",
  "These pathways collapsed to ", 
  nrow(parent_enriched_pathways), 
  " parent Reactome pathway terms."
)

message
```

The mean bill length of penguins in the dataset is “r round(mean(na.omit(penguins$bill_length_mm)),2)”mm.

### Tidy up the environment and R memory
```{r}
# rm(list = ls())
# gc()
```
