---
title: "Negative Controls for Differential Gene Expression Analysis and Gene Set Enrichment Analysis"
author: "Kieran Redpath"
format: html
editor: source
theme: cosmo
---

## Summary

-   Reloads processed data from 0_Setup.qmd.
-   Carries out negative controls for GSEA.
-   Saves output as a .rds file for reference.

### Start with a clean environment

```{r}
rm(list = ls())
gc()
```

### Load packages

```{r}
#| results: hide
#| warning: false
#| message: false
#| echo: false

library(ggplot2)
library(limma)
library(edgeR)
library(data.table)
library(magrittr)
library(CePa)
library(tidyverse)
library(ggbeeswarm)
library(dplyr)
library(ReactomePA)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(reactome.db)
library(goseq)
library(gplots)
library(devtools)
library(viridis)
library(openxlsx)
library(fgsea)
library(evaluate)
library(gridExtra)
library(synergyfinder)
library(cowplot)

# Set the random seed so results are reproducible.
set.seed(42)
```

### Load pre-processed data

```{r}
# GDSC DOI response data for cell lines used in DE analysis.
sort_DOI_GDSC <- readRDS("data/sort_DOI_GDSC.rds")

# CCLE normalized gene expression data for cell lines used in DE analysis.
sort_voom_CCLE <- readRDS("data/sort_voom_CCLE.rds")

# Tool for converting gene names without calling AnnotationDbi::select() every time.
gene_tool <- readRDS("data/gene_tool.rds")

# Tool for indexing cell lines to tissues.
tissue_tool <- readRDS("data/tissue_tool.rds")
```

### Examine the data

```{r}
# Check that all column names are identical between the CCLE and GDSC datasets.
sum(colnames(sort_voom_CCLE)==sort_DOI_GDSC$CCLE_Name)
# What types of data are included in the processed GDSC dataset?
names(sort_DOI_GDSC)
```

### Randomise row names (genes) for negative controls

-   Carries out ten iterations of:
    -   Randomising rownames.

    -   Carrying out differential expression analysis on randomised genes.

    -   Calculating empirical p-values for enriched genes with 500,000 iterations of Reactome pathway enrichment analysis using fgsea.

    -   Stores these results in a list called results, where each element is one of the ten iterations.
-   Calculates the number of adjusted p-values < 0.05 in any element of results.

```{r}
#| results: hide
#| warning: false
#| message: false

# Define number of iterations to carry out.
iterations <- 10
# Create an object to store results in.
results <- list()

# Carry out row name randomisation a number of times equal to iterations.
for (i in 1:iterations) {
  cat("Iteration", i, "of", iterations, "\n")
  # Create a randomised list of the gene names and apply them to the matrix.
  sort_voom_CCLE_negative <- sort_voom_CCLE
  rownames(sort_voom_CCLE_negative) <- sample(rownames(sort_voom_CCLE_negative))
  
  # Generate the design matrix for limma, defining groups based on response to DOI.
  design <- model.matrix(~cut(sort_DOI_GDSC$AUC,
                              breaks = quantile(sort_DOI_GDSC$AUC, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
                              labels = c("Bottom", "Middle", "Top"),
                              include.lowest = TRUE))
  colnames(design) <- c("Mean", "BottomVsMiddle", "BottomVsTop")
  
  # Fit the linear model.
  fit <- lmFit(sort_voom_CCLE_negative, design) %>% eBayes(.)
  # Extract top table.
  top_table <- topTable(fit, coef = "BottomVsTop", adjust = "BH", n = nrow(sort_voom_CCLE_negative)) %>% arrange(adj.P.Val)
  
  # Format genes for Reactome pathway analysis, converting to Entrez IDs for fgsea, while handling missing and duplicate Entrez IDs.
  ranked_genes <- top_table %>%
    rownames_to_column(var = "ENSEMBL") %>% # Convert rownames to a column.
    arrange(adj.P.Val) %>% # Ensure that the top_table is sorted by BH-adjusted p-value.
    left_join(gene_tool, by = c("ENSEMBL" = "ENSEMBL")) %>% # Add information for Entrez ID and gene symbol.
    dplyr::filter(!is.na(ENTREZID)) %>% # Remove genes without Entrez gene IDs.
    distinct(ENTREZID, .keep_all = TRUE) %>% # In case of duplicates, keep only the first occurrence of each Entrez ID (i.e. the one with the lowest BH-adjusted p-value).
    dplyr::select(ENTREZID, t) %>% # Keep only Entrez IDs and t-statistic.
    arrange(t) %>% # Arrange by t-statistic for GSEA.
    deframe() # Convert to named vector.
  
  # Carry out GSEA.
  # Find Reactome pathways that these genes are present in.
  all_pathways <- ranked_genes %>% names(.) %>% reactomePathways(.)
  # 1:many mapping reflects that many genes map to multiple Reactome pathways.
  # Find enriched Reactome pathways.
  results[[i]] <- fgsea(pathways = all_pathways, stats = ranked_genes, maxSize = 500, nperm = 500000) %>% arrange(padj)
    # Count how many rows have padj < 0.05.
  sig_count <- sum(results[[i]]$padj < 0.05, na.rm = TRUE)
  
  # Print iteration and count
  cat("Iteration", i, ":", sig_count, "significant results (padj < 0.05)\n")
}

# Tidy up the environment.
# rm(list = c("iterations", "i"))
```

### Report any significantly enriched Reactome pathways across all negative controls

```{r}
for (i in 1:iterations) {
  # Count how many rows have padj < 0.05.
  sig_count <- sum(results[[i]]$padj < 0.05, na.rm = TRUE)
  
  # Print iteration and count.
  cat("Iteration", i, ":", sig_count, "significant results (padj < 0.05)\n")
}

# Tidy up the environment.
rm(list = c("sig_count"))
```

### Save pre-processed objects as .rds files

```{r}
# Negative control results.
saveRDS(results, file = "data/negative_control_results.rds")
```

### Tidy up the environment and R memory

```{r}
rm(list = ls())
gc()
```
