---
title: "Package_Control_renv"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Get renv working with R 3.6.2
```{r}
install.packages("renv", version = "0.9.2")
```

### Install packages unlikely to be troublemakers but still required (most recent version at 10/10/2022)
```{r}
install.packages("data.table")

install.packages("magrittr")

install.packages("ggplot2")

install.packages("tidyverse")

install.packages("ggbeeswarm")

install.packages("dplyr")

install.packages("gplots")

install.packages("knitr")

install.packages("kableExtra")

install.packages("devtools")

install.packages("viridis")

install.packages("evaluate")

install.packages("tibble")
```

### BiocManager itself isn't causing any problems, but a specific version needs to be installed in order to set package versions
```{r}
install.packages("BiocManager")

# Revert Bioconductor to version 3.9
BiocManager::install(version = '3.9', update = FALSE, force = TRUE)

# Try version 3.10
# BiocManager::install(version = '3.10', update = FALSE, force = TRUE)
```

### Install older version of CePa
```{r}
BiocManager::install("Rgraphviz", version = "3.9", update = FALSE, force = TRUE)
# BiocManager::install("Rgraphviz", version = "3.10", update = FALSE, force = TRUE)
require(devtools)
install_version("CePa", version = "0.7.0", repos = "http://cran.us.r-project.org")
```

### Install packages that might be responsible for causing problems (version as at May 20, 2021, Bioconductor Release 3.9 for R Version 3.6.2)
```{r}
 ### A few pre-requisites to make goseq work properly
install.packages("~/Misc/RCurl_1.98-1.8.zip", repos = NULL, type = "win.binary")
BiocManager::install(c("limma", "edgeR", "org.Hs.eg.db", "ReactomePA", "reactome.db", "geneLenDataBase", "goseq"), version = "3.9", update = FALSE, force = TRUE)

# Bioconductor release 3.10
BiocManager::install(c("limma", "edgeR", "org.Hs.eg.db", "ReactomePA", "reactome.db", "geneLenDataBase", "goseq"), version = "3.10", update = FALSE, force = TRUE)
```

### Load packages
```{r}

```

