---
title: "Setup"
author: "Kieran Redpath"
format: html
editor: source
theme: cosmo
---

## Summary

-   Loads and normalizes gene expression data from the CCLE.
-   Loads and processes drug response data from the GDSC.
-   Filters these data to the relevant information for this project.
-   Saves these outputs as .rds files for later use.

### Start with a clean environment

```{r}
rm(list = ls())
gc()
```

### Load packages

```{r}
#| results: hide
#| warning: false
#| message: false
#| echo: false

library(ggplot2)
library(limma)
library(edgeR)
library(data.table)
library(magrittr)
library(CePa)
library(tidyverse)
library(ggbeeswarm)
library(dplyr)
library(ReactomePA)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(reactome.db)
library(goseq)
library(gplots)
library(devtools)
library(viridis)
library(openxlsx)
library(fgsea)
library(evaluate)
library(gridExtra)
library(synergyfinder)
library(cowplot)

# Set the random seed so results are reproducible.
set.seed(42)
```

### Define the drug of interest

```{r}
DOI <- "Dasatinib"
```

### Load CCLE gene expression data

```{r}
# Load Homo sapiens gene data (including Entrez ID's).
hs <- org.Hs.eg.db

# Load expression data from the CCLE, available here: https://depmap.org/portal/data_page/?tab=allData
CCLE <- read.gct("data/raw/CCLE_RNAseq_genes_counts_20180929.gct")
# Remove X's introduced by R for cell lines that start with a number, but not those that don't.
colnames(CCLE) <- sub("^X(\\d+)", "\\1", colnames(CCLE))
# Convert gene names to stable Ensembl IDs by removing version numbers.
rownames(CCLE) <- sub("\\..*", "", rownames(CCLE))
# Ensure that cell line names are all uppercase for consistency.
colnames(CCLE) <- toupper(colnames(CCLE))
# Normalise CCLE counts using the limma voom method.
dge <- DGEList(counts = CCLE)
dge <- calcNormFactors(dge)
voom_dge <- voom(dge, plot = TRUE)
voom_CCLE <- voom_dge$E

# Tidy up the environment.
rm(list = c("dge", "voom_dge"))
```

### Load GDSC drug response data

```{r}
# Load the GDSC dataset, current release available here: www.cancerrxgene.org/downloads/bulk_download
# Archived releases available via ftp: ftp.sanger.ac.uk/project/cancerrxgene/releases/
GDSC <- read.xlsx("data/raw/GDSC2_fitted_dose_response_27Oct23.xlsx")

# Convert column names to uppercase for consistency with CCLE.
GDSC$CELL_LINE_NAME <- toupper(GDSC$CELL_LINE_NAME)

# Resolve discrepancies between IC50 values for drugs tested across multiple dose response curves.
GDSC <- GDSC %>%
  group_by(CELL_LINE_NAME, DRUG_NAME) %>%
  mutate(MEAN_LN_IC50 = mean(LN_IC50, na.rm = TRUE)) %>%
  ungroup()
```

### Format data

```{r}
#| warning: false

# Standardize CCLE and GDSC cell line names, resolving a naming discrepancy between between cellosaurus cell lines www.cellosaurus.org/CVCL_3174 and www.cellosaurus.org/CVCL_1774, as well as https://www.cellosaurus.org/CVCL_S641 and https://www.cellosaurus.org/CVCL_1330
# CCLE:
colnames(voom_CCLE)[colnames(voom_CCLE) == "TT_OESOPHAGUS"] <- "T-T_OESOPHAGUS"
colnames(voom_CCLE)[colnames(voom_CCLE) == "KMH2_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE"] <- "KM-H2_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE"
# Separate cell line and tissue identifiers for later indexing.
split_cell_line_tissues <- colnames(voom_CCLE) %>% strsplit(., "_")
colnames(voom_CCLE) <- strsplit(colnames(voom_CCLE), "_") %>% 
  lapply(., function(x) x[1]) %>% 
  unlist()
# GDSC:
GDSC <- GDSC %>%
  mutate(CELL_LINE_NAME = case_when(
    CELL_LINE_NAME == "T-T"    ~ "T-T",
    CELL_LINE_NAME == "KM-H2"  ~ "KM-H2",
    CELL_LINE_NAME == "KMH-2"  ~ "KMH-2",
    TRUE ~ gsub("-", "", CELL_LINE_NAME, fixed = TRUE)
  ))

# Extract DOI response data from GDSC.
DOI_GDSC <- filter(GDSC, DRUG_NAME == DOI)

# Filter datasets for common cell lines.
common_lines <- intersect(DOI_GDSC$CELL_LINE_NAME,colnames(voom_CCLE))
match_voom_CCLE <- match(common_lines, colnames(voom_CCLE))
common_voom_CCLE <- voom_CCLE[ , na.omit(match_voom_CCLE)]
match_DOI_GDSC <- match(common_lines, DOI_GDSC$CELL_LINE_NAME)
common_DOI_GDSC <- DOI_GDSC[na.omit(match_DOI_GDSC),]
colnames(common_DOI_GDSC)[5] <- "CCLE_Name"

# Tidy up the environment.
rm(list = c("common_lines", "match_voom_CCLE", "match_DOI_GDSC"))
```

### Filter data

```{r}
# Filter CCLE and GDSC data for BRCA and STAD cell lines with data for DOI.
# GDSC (also adds a column for CDH1 expression data):
sort_DOI_GDSC <- common_DOI_GDSC %>%
  filter(TCGA_DESC %in% c("BRCA", "STAD"))
sort_DOI_GDSC$CDH1 <- common_voom_CCLE %>%
  as.data.frame() %>%
  dplyr::select(any_of(sort_DOI_GDSC$CCLE_Name)) %>%
  dplyr::slice(match("ENSG00000039068", rownames(.))) %>%
  unlist(use.names = FALSE) # flatten to a vector.

# CCLE:
sort_voom_CCLE <- common_voom_CCLE %>%
  as.data.frame() %>%
  dplyr::select(any_of(sort_DOI_GDSC$CCLE_Name)) %>% 
  as.matrix()

# Tidy up the environment.
rm(list = c("common_voom_CCLE", "common_DOI_GDSC"))
```

### Index cell lines to tissues

```{r}
# Create a tool to find tissue type based on cell line name or vice versa.
tissue_tool <- data.frame(
  Cell_Line = sapply(split_cell_line_tissues, `[[`, 1),
  Tissue_Type = sapply(split_cell_line_tissues, function(x) paste(x[-1], collapse = "_")),
  stringsAsFactors = FALSE)
tissue_tool <- tissue_tool %>% mutate(
  Tissue_Type = Tissue_Type %>%
    str_remove_all("c") %>%
    str_replace_all("[[:punct:]]", "_"))

# Create another tool for the same purpose, but only including breast and stomach cell lines with DOI response data.
sort_tissue_tool <- tissue_tool %>% filter(Tissue_Type %in% c("BREAST", "STOMACH")) # Cell lines from CCLE.
common_lines <- intersect(colnames(sort_voom_CCLE), sort_tissue_tool$Cell_Line) # Cell lines in GDSC with DOI data.
sort_tissue_tool <- sort_tissue_tool %>% # Overlap between the two.
  filter(Cell_Line %in% common_lines) %>%
  arrange(factor(Cell_Line, levels = common_lines))

# Tidy up the environment.
rm(list = c("split_cell_line_tissues", "common_lines"))
```

### Index alternative identifiers for all genes in the dataset

```{r}
# Create the gene name reference tool.
gene_tool <- rownames(CCLE) %>%
  AnnotationDbi::select(hs,
                        keys = .,
                        columns = c("ENSEMBL", "ENTREZID", "SYMBOL"),
                        keytype = "ENSEMBL")
```

### Save pre-processed objects as .rds files

```{r}
# GDSC.
saveRDS(GDSC, file = "data/GDSC.rds")

# DOI_GDSC.
saveRDS(DOI_GDSC, file = "data/DOI_GDSC.rds")

# sort_DOI_GDSC.
saveRDS(sort_DOI_GDSC, file = "data/sort_DOI_GDSC.rds")

# voom_CCLE.
saveRDS(voom_CCLE, file = "data/voom_CCLE.rds")

# sort_voom_CCLE.
saveRDS(sort_voom_CCLE, file = "data/sort_voom_CCLE.rds")

# tissue_tool.
saveRDS(tissue_tool, file = "data/tissue_tool.rds")

# sort_tissue_tool.
saveRDS(sort_tissue_tool, file = "data/sort_tissue_tool.rds")

# gene_tool.
saveRDS(gene_tool, file = "data/gene_tool.rds")
```

### Tidy up the environment and R memory

```{r}
rm(list = ls())
gc()
```
