---
title: "Differential Gene Expression Analysis"
author: "Kieran Redpath"
format: html
editor: source
theme: cosmo
---

## Summary

-   Carries out differential expression analysis between BRCA and STAD cell lines with the top and bottom tertiles for DOI response.
-   Saves these outputs as .rds files for later use.

### Start with a clean environment

```{r}
rm(list = ls())
gc()
```

### Load packages

```{r}
#| results: hide
#| warning: false
#| message: false
#| echo: false

library(ggplot2)
library(limma)
library(edgeR)
library(data.table)
library(magrittr)
library(CePa)
library(tidyverse)
library(ggbeeswarm)
library(dplyr)
library(ReactomePA)
library(org.Hs.eg.db)
library(reactome.db)
library(goseq)
library(gplots)
source('data/raw/heatmap-mik.R')
library(knitr)
library(kableExtra)
library(devtools)
library(viridis)
library(openxlsx)

# Set the random seed so results are reproducible.
set.seed(42)
```

### Load pre-processed data

```{r}
# Tool for converting gene names without calling AnnotationDbi::select() every time.
gene_tool <- readRDS("data/gene_tool.rds")

# GDSC DOI response data for cell lines used in DE analysis.
sort_DOI_GDSC <- readRDS("data/sort_DOI_GDSC.rds")

# CCLE normalized gene expression data for cell lines used in DE analysis.
sort_voom_CCLE <- readRDS("data/sort_voom_CCLE.rds")
```

### Examine the data

```{r}
# Check that all column names are identical between the CCLE and GDSC datasets
sum(colnames(sort_voom_CCLE)==sort_DOI_GDSC$CCLE_Name)
# What types of data are included in the processed GDSC dataset?
names(sort_DOI_GDSC)
```

### Differential expression analysis

```{r}
# Create groups for comparison, based on drug response to DOI.
group <- cut(
  sort_DOI_GDSC$AUC,
  breaks = quantile(sort_DOI_GDSC$AUC, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
  labels = c("Bottom", "Middle", "Top"),
  include.lowest = TRUE
)

# How many cell lines are in each group?
table(group)
# Visualize the ranges of drug response values in each group.
ggplot(sort_DOI_GDSC, aes(x = group, y = AUC)) +
  geom_boxplot() +
  labs(title = "AUC Distribution by Group",
       x = "Group", y = "AUC") +
  theme_minimal()
# Create the design matrix.
design <- model.matrix(~group); colnames(design) <- c("Mean", "BottomVsMiddle", "BottomVsTop")
# Fit the linear model.
fit <- lmFit(sort_voom_CCLE, design) %>% eBayes(.)
# Create the top table of DE genes, and ensure that it's sorted by adjusted p-value.
top_table <- topTable(fit, coef = "BottomVsTop", adjust = "BH", n = nrow(sort_voom_CCLE)) %>%
  arrange(adj.P.Val)

# Tidy up the environment.
rm(list = c("group", "design", "fit"))
```

### Explore the differential expression data

```{r}
# Plot the counts distribution of the gene with the smallest adjusted p-value across these cell lines.
sort_voom_CCLE[,1] %>% density(.) %>% plot(.)
# How many genes are significantly differentially expressed at BH-adjusted α = 0.05?
sum(top_table$adj.P.Val < 0.05)

# Plot fold change in AUC against adjusted p-value.
# Create new columns in top_table for plotting.
top_table$logAdjP <- -log10(top_table$adj.P.Val)
top_table$significant <- (top_table$adj.P.Val < 0.05)
# Define the adjusted p-value threshold line.
adj_p_thresh <- -log10(0.05)
# Plot differential expression for diagnosis.
volcano <- ggplot(top_table, aes(x = logFC, y = logAdjP, legend_position = "none")) +
  geom_point(aes(color = significant), size = 1) +
  scale_color_manual(values = c("black", "red")) +
  geom_hline(yintercept = adj_p_thresh, linetype = "dashed", color = "blue") +
  labs(title = "Volcano Plot (Adjusted P-Values)", x = "log2 Fold Change", y = "-log10(Adjusted P-Value)") +
  theme_minimal() +
  theme(legend.position = "none")
# Print.
volcano
# Save.
ggsave("1_output/volcano.png", plot = volcano, width = 3000, height = 2000, units = "px")

# Tidy up the environment.
rm(list = c("adj_p_thresh"))
```

### Save processed objects

```{r}
# Remove plotting columns from top_table and save as .rds.
top_table <- top_table %>% dplyr::select(-logAdjP, -significant)
# top_table.
saveRDS(top_table, file = "1_output/top_table.rds")

# Create a copy of top_table for publication.
top_table_pub <- top_table

# Convert Ensembl ID rownames to a column.
top_table_pub <- rownames_to_column(top_table_pub, "ENSEMBL")

# Add a column for gene symbols converted from Ensembl ID's.
top_table_pub <- top_table_pub %>% 
left_join(gene_tool, by = c("ENSEMBL" = "ENSEMBL"))

# Keep specified columns and reorder.
top_table_pub <- top_table_pub %>%
  dplyr::select(ENSEMBL, SYMBOL, adj.P.Val, t, AveExpr, logFC)

# Keep only genes that are significant at BH-adjusted α < 0.05.
top_table_pub <- top_table_pub %>% 
  filter(adj.P.Val < 0.05)

# Rename columns.
top_table_pub <- dplyr::rename(top_table_pub, "Ensembl ID" = "ENSEMBL", "Gene Symbol" = "SYMBOL", "BH-Adjusted P-Value" = "adj.P.Val", "T-Statistic" = "t", "Average Expression" = "AveExpr", "Log2 Fold Change in Expression" = "logFC")

# Round all numeric columns to 4 decimal places.
top_table_pub[] <- lapply(top_table_pub, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

# Inspect the top 10 differentially expressed genes.
head(top_table_pub, 10)

# Save table.
write.csv(top_table_pub, file = "1_output/top_table_pub.csv")
```

### Tidy up the environment and R memory

```{r}
rm(list = ls())
gc()
```
