---
title: "Differential Gene Expression Analysis"
author: "Kieran Redpath"
format: html
editor: source
theme: cosmo
---

## Summary
+ Reloads processed data from 0_Setup.qmd
+ Carries out differential expression analysis between BRCA and STAD cell lines with the top and bottom tertiles for dasatinib response.

### Load packages
```{r}
#| results: hide
#| warning: false
#| message: false
#| echo: false

library(limma)
library(edgeR)
library(data.table)
library(magrittr)
library(ggplot2)
library(CePa)
library(tidyverse)
library(ggbeeswarm)
library(dplyr)
library(ReactomePA)
library(org.Hs.eg.db)
library(reactome.db)
library(goseq)
library(gplots)
source('data/heatmap-mik.R')
library(knitr)
library(kableExtra)
library(devtools)
library(viridis)
library(openxlsx)
set.seed(42)
```

### Load pre-processed data
```{r}
# GDSC2 dasatinib response data for cell lines used in DE analysis.
sort_dasatinib_GDSC2 <- readRDS("data/sort_dasatinib_GDSC2.rds")

# CCLE normalized gene expression data for cell lines used in DE analysis.
sort_voom_CCLE <- readRDS("data/sort_voom_CCLE.rds")
```

### Examine the data
```{r}
# Check that all column names are identical between the CCLE and GDSC datasets
sum(colnames(sort_voom_CCLE)==sort_dasatinib_GDSC2$CCLE_Name)
# What types of data are shown in the GDSC2 dataset?
names(sort_dasatinib_GDSC2)
```

### Differential expression analysis
```{r}
# Create groups for comparison, based on drug response to dasatinib
group <- cut(
  sort_dasatinib_GDSC2$LN_IC50,
  breaks = quantile(sort_dasatinib_GDSC2$LN_IC50, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
  labels = c("Bottom", "Middle", "Top"),
  include.lowest = TRUE
)

# How many cell lines are in each group?
table(group)
# Visualize the ranges of drug response values in each group.
ggplot(sort_dasatinib_GDSC2, aes(x = group, y = LN_IC50)) +
  geom_boxplot() +
  labs(title = "IC50 Distribution by Group",
       x = "Group", y = "IC50") +
  theme_minimal()
# Create the design matrix.
design <- model.matrix(~group); colnames(design) <- c("Mean", "BottomVsMiddle", "BottomVsTop")
# Fit the linear model.
fit <- lmFit(sort_voom_CCLE, design) %>% eBayes(.)
# Create the top table of DE genes, and ensure that it's sorted by adjusted p-value.
top_table <- topTable(fit, coef = "BottomVsTop", adjust = "BH", n = nrow(sort_voom_CCLE)) %>%
  arrange(adj.P.Val)

# Tidy up the environment.
rm(list = c("group", "design", "fit"))
```

### Explore the differential expression data
```{r}
# Plot the counts distribution of the gene with the smallest adjusted p-value across these cell lines.
sort_voom_CCLE[,1] %>% density(.) %>% plot(.)
# How many genes are significantly differentially expressed at α = 0.05?
sum(top_table$adj.P.Val < 0.05)

# Plot fold change in AUC against adjusted p-value.
# Create new columns in top_table for plotting.
top_table$logAdjP <- -log10(top_table$adj.P.Val)
top_table$highlight <- (top_table$adj.P.Val < 0.05) & (abs(top_table$logFC) > 1)
# Define the adjusted p-value threshold line.
adj_p_thresh <- -log10(0.05)
# Plot.
ggplot(top_table, aes(x = logFC, y = logAdjP)) +
  geom_point(aes(color = highlight), size = 1) +
  scale_color_manual(values = c("black", "red")) +
  geom_hline(yintercept = adj_p_thresh, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  labs(title = "Volcano Plot (Adjusted P-values)", x = "log2 Fold Change", y = "-log10(Adjusted P-value)") +
  theme_minimal()

# Tidy up the environment.
rm(list = c("adj_p_thresh"))
```

### Save processed objects as .rds files
```{r}
# top_table
saveRDS(top_table, file = "1_output/top_table.rds")
```

### Tidy up the environment and R memory
```{r}
rm(list = ls())
gc()
```
