---
title: "Dasatinib_fGSEA_Analysis"
author: "Kieran Redpath"
date: "19 March 2021"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Packages
```{r, results='hide', warning=FALSE, message=FALSE}
library(CePa)
library(data.table)
library(dplyr)
library(edgeR)
library(fgsea)
library(ggplot2)
library(limma)
library(reactome.db)
library(org.Hs.eg.db)
set.seed(42)
```

### Load and process the data
```{r}
hs <- (org.Hs.eg.db)

# Load raw data, and normalise raw expression data.
ccleExpData <- read.gct('Data/CCLE_RNAseq_genes_counts_20180929.gct')
# Remove X's from cell lines that start with a number.
colnames(ccleExpData) <- sub("^X", "", colnames(ccleExpData))
# Normalise with the edgeR and limma packages.
dge <- DGEList(counts = ccleExpData)
dge <- calcNormFactors(dge)
dge_voom <- voom(dge, plot = FALSE)
expDat <- dge_voom$E
expDatRaw <- expDat

# GDSC2 Data for differential expression analysis.
GDSC2 <- fread('Data/GDSC2_fitted_dose_response_25Feb20.csv', sep= ';')
dim(GDSC2)
dasatinib <- GDSC2 %>% filter(., DRUG_NAME=="Dasatinib")
colnames(expDat) <- strsplit(colnames(expDat), "_") %>% 
  lapply(., function(x) x[1]) %>% 
  unlist()

# Filters for cell lines that are shared between the GDSC and CCLE datasets.
dasatinib$CELL_LINE_NAME <- gsub("-","",dasatinib$CELL_LINE_NAME, fixed=TRUE)
commonSamples <- intersect(dasatinib$CELL_LINE_NAME,colnames(expDat))
expDat_match <- match(commonSamples, colnames(expDat))
expDat_sort <- expDat[ , na.omit(expDat_match)]
dasatinib_match <- match(commonSamples, dasatinib$CELL_LINE_NAME)
dasatinib_sort <- dasatinib[na.omit(dasatinib_match),]
colnames(dasatinib_sort)[5] <- "CCLE_Name"
# Down to 472 cell lines

# Filters dasatinib data for the two tissues of interest, and removes the middle 33% of samples (based on AUC), so that we're only looking at the extremes. Then filters "expDat_sort" so it's also only these cell lines, those that have data for the drug of interest, and CDH1 expression data.
dasatinib_sort <- dasatinib_sort %>% filter(., TCGA_DESC=="BRCA" | TCGA_DESC=="STAD")  
dasatinib_sort <- dasatinib_sort[ dasatinib_sort$AUC < quantile(dasatinib_sort$AUC , 0.33) | dasatinib_sort$AUC > quantile(dasatinib_sort$AUC, 0.66), ]
dasatinib_sort$CDH1 <- expDat_sort %>% as.data.frame() %>% dplyr::select(., which(colnames(.) %in% dasatinib_sort$CCLE_Name)) %>%  .["ENSG00000039068.14",] %>% as.vector() %>% t()
expDat_sort <- expDat_sort %>% subset(., select=which(colnames(expDat_sort) %in% dasatinib_sort$CCLE_Name))


# Gene label tool to convert gene ID's.
GeneLabelTool <- strsplit(rownames(expDatRaw), '\\.') %>% 
  lapply(., function(x) x[1]) %>% 
  unlist()
GeneLabelTool <- AnnotationDbi::select(hs,
                                       keys = GeneLabelTool,
                                       columns = c("ENSEMBL", "ENTREZID", "SYMBOL"),
                                       keytype = "ENSEMBL")
GeneLabelTool <- na.omit(GeneLabelTool)
```
### Manage existing data
```{r}
# Look at the existing data
sum(colnames(expDat_sort)==dasatinib_sort$CCLE_Name)
names(dasatinib_sort)

# Fits the expression data to a linear model.
group2 <- ifelse(dasatinib_sort$AUC > median(dasatinib_sort$AUC), "High", "Low")
table(group2)
boxplot(dasatinib_sort$AUC ~ group2)
design2 = model.matrix(~group2);
design2 %>% head()
colnames(design2) = c("Mean", "HighVsLow")
fit2 = lmFit(expDat_sort, design2)
fit2 = eBayes(fit2)
tt2 = topTable(fit2, coef="HighVsLow", adjust="BH",n=nrow(expDat_sort))
options(digits=4)
```


### Carry out GSEA to find significant reactome pathways
```{r}
# Format the gene list for reactome pathway analysis.
de <- data.table(tt2, keep.rownames = TRUE)
ranks <- de[order(t), list(rn, t)]

# Make sure ranks$rn matches the format of GeneLabelTool (no alternate transcripts).
ranks$rn <- strsplit(ranks$rn, '\\.') %>% 
  lapply(., function(x) x[1]) %>% 
  unlist()

# Convert gene list from ENSG to Entrez ID's.
ranks$rn <- match(ranks$rn, GeneLabelTool$ENSEMBL) %>% 
    GeneLabelTool$ENTREZID[.]
# Remove NA's from gene list.
ranks <- na.omit(ranks)

# Find reactome pathways.
fgseapathways <- reactomePathways(ranks$rn)


# convert the gene list into the right format for analysis.
ranks2 <- ranks$t
names(ranks2) <- ranks$rn
ranks2 <- ranks2[!duplicated(names(ranks2))]

expDat_Dup <- expDat[ , !duplicated(colnames(expDat))]  
  
# Carry out gsea analysis.
fgseaRes <- fgsea(fgseapathways, ranks2, maxSize=500)
fgseaRes <- fgseaRes[order(padj),]
```

### Compare GSEA pathways to goseq pathways
```{r}
# Load goseqPathways.
goseqPathways <- read.csv("Data/goseq_Das_Pathways.csv", row.names = 1)

# Compare the top 100 pathways between goseq and gsea analysis.
# intersect(fgseaRes$pathway[1:100], goseqPathways$Pathway[1:100])
intersect(fgseaRes$pathway[1:100], goseqPathways$Pathway[1:100]) %>% length(.)
# 39 of which overlap (with current lists).

# Returns the position in the second argument of the values in the first argument.
# match(goseqPathways$Pathway[1:100], fgseaRes$pathway[1:100])

# Compare all of the pathways shared between goseq and gsea analysis.
# intersect(fgseaRes$pathway, goseqPathways$Pathway)
intersect(fgseaRes$pathway, goseqPathways$Pathway) %>% length(.)
# 349 of which overlap (with current lists).
```

There isn't the craziest amount of overlap (especially in the top 100), especially when you you look at the positions of the goseqPathways in the fgseaRes pathways.

### Look at the overlap between significant pathways
```{r}
# Subset to find significant pathways in each pathway list.
sigfgseaRes <- fgseaRes[fgseaRes$padj < 0.05, ]
nrow(sigfgseaRes)
siggoseqPathways <- goseqPathways[goseqPathways$adjP < 0.05, ]
nrow(siggoseqPathways)

intersect(sigfgseaRes$pathway, siggoseqPathways$Pathway)
intersect(sigfgseaRes$pathway, siggoseqPathways$Pathway) %>% length(.)
# 51 of these pathways overlap (with all  SigGenes, where there's only 76 from goseq anyway).
# 27 of these pathways overlap (with only 2000 SigGenes, where there's only 28 from goseq anyway).
```

### Try collapsing pathways so there's less overlap between them for analysis.
```{r, warning=FALSE}
# Collapse Pathways
collapsedPathways <- collapsePathways(fgseaRes[padj < 0.05],
                                      fgseapathways, ranks2)
# Analyse these a bit.
mainPathways <- fgseaRes[pathway %in% collapsedPathways$mainPathways][
order(-NES), pathway]
# Cool.
```

### Plot the first couple of pathways as GSEA enrichment plots.
```{r}
# Plot the first pathway.
plotEnrichment(fgseapathways[["Assembly of collagen fibrils and other multimeric structures"]], ranks2)
# Plot the second pathway.
plotEnrichment(fgseapathways[["Cell surface interactions at the vascular wall"]], ranks2)
# Plot the third pathway.
plotEnrichment(fgseapathways[["Collagen biosynthesis and modifying enzymes"]], ranks2)
# Plot the very last pathway for comparison.
plotEnrichment(fgseapathways[["Manipulation of host energy metabolism"]], ranks2)

# Plot the table - not really sure what's happening here or what I'm supposed to be seeing
topPathways <- fgseaRes[head(order(padj), n = 20)][order(NES), pathway]
pdf(file = "Dasatinib_fGSEA_Analysis_Output/Other/fgseaOutput.pdf", width = 11, height = 25)
plotGseaTable(fgseapathways[topPathways], ranks2, fgseaRes, gseaParam=0.5, colwidths = c(5, 3, 0.8, 1.2, 1.2), render = TRUE)
dev.off()
```



