---
title: "Gene Set Enrichment Analysis"
author: "Kieran Redpath"
format: html
editor: source
theme: cosmo
---


## Summary
+ Reloads processed data from 0_Setup.qmd and 1_Differential_Expression_Analysis.qmd
+ Carries out gene set enrichment analysis on differentially expressed gene list, ranked by limma-moderated t-statistic.

### Load packages
```{r}
#| results: hide
#| warning: false
#| message: false
#| echo: false

library(CePa)
library(data.table)
library(dplyr)
library(edgeR)
library(fgsea)
library(ggplot2)
library(limma)
library(reactome.db)
set.seed(42)
```

### Load pre-processed data
```{r}
# Tool for converting gene names without calling AnnotationDbi::select() every time.
gene_tool <- readRDS("data/gene_tool.rds")

# Tool for indexing cell lines to tissues.
tissue_tool <- readRDS("data/tissue_tool.rds")

# Top table from differential expression analysis.
top_table <- readRDS("1_output/top_table.rds")
```

### Rank genes by limma-moderated t-statistic
```{r}
# Format genes for Reactome pathway analysis, converting from Ensembl to Entrez ID's, and removing any that don't have Entrez ID's.
ranked_genes <- top_table %>%
  rownames_to_column(var = "rn") %>%
  arrange(t) %>%
  left_join(gene_tool, by = c("rn" = "ENSEMBL")) %>%
  dplyr::filter(!is.na(ENTREZID)) %>% # Remove genes without Entrez gene IDs for fgsea.
  dplyr::select(ENTREZID, t) %>%
  deframe()
```

### Carry out gene set enrichment analysis
```{r}
#| warning: false

# Find Reactome pathways that these genes are present in.
all_pathways <- ranked_genes %>% names(.) %>% reactomePathways(.)

# Carry out gsea analysis.
gsea_results <- fgsea(pathways = all_pathways, stats = ranked_genes, maxSize = 500) %>% 
  arrange(padj) # Dataset contains some duplicate Entrez gene IDs.

# Identify pathways that are significantly enriched at a given BH-adjusted alpha.
enriched_pathways <- gsea_results %>% dplyr::filter(padj < 0.05) %>% 
  arrange(padj)

# Identify shared parent pathways among significantly enriched pathways.
parent_pathway_mapping <- collapsePathways(enriched_pathways, all_pathways, ranked_genes)
parent_enriched_pathways <- enriched_pathways %>%
  dplyr::filter(pathway %in% parent_pathway_mapping$mainPathways) %>%
  arrange(padj)

# Tidy up the environment.
rm(list = c("ranked_genes", "all_pathways", "gsea_results", "parent_pathway_mapping"))
```

### Save processed objects as .rds files
```{r}
# enriched_pathways
saveRDS(enriched_pathways, file = "2_output/enriched_pathways.rds")

# parent_enriched_pathways
saveRDS(parent_enriched_pathways, file = "2_output/parent_enriched_pathways.rds")
```

### Tidy up the environment and R memory
```{r}
# rm(list = ls())
# gc()
```
