---
title: "Gene Set Enrichment Analysis"
author: "Kieran Redpath"
format: html
editor: source
theme: cosmo
---

## Summary

-   Reloads processed data from 0_Setup.qmd and 1_Differential_Expression_Analysis.qmd
-   Carries out gene set enrichment analysis for Reactome pathways on differentially expressed gene list, ranked by limma-moderated t-statistic.
-   Collapses Reactome pathway terms to shared parent pathway terms.
-   Saves these outputs as .rds files for later use.

### Start with a clean environment

```{r}
rm(list = ls())
gc()
```

### Load packages

```{r}
#| results: hide
#| warning: false
#| message: false
#| echo: false

library(ggplot2)
library(limma)
library(edgeR)
library(data.table)
library(magrittr)
library(CePa)
library(tidyverse)
library(ggbeeswarm)
library(dplyr)
library(ReactomePA)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(reactome.db)
library(goseq)
library(gplots)
library(devtools)
library(viridis)
library(openxlsx)
library(fgsea)
library(evaluate)
library(gridExtra)
library(synergyfinder)
library(cowplot)

# Set the random seed so results are reproducible.
set.seed(42)
```

### Load pre-processed data

```{r}
# Tool for converting gene names without calling AnnotationDbi::select() every time.
gene_tool <- readRDS("data/gene_tool.rds")

# Tool for indexing cell lines to tissues.
tissue_tool <- readRDS("data/tissue_tool.rds")

# Top table from differential expression analysis.
top_table <- readRDS("1_output/top_table.rds")
```

### Rank genes by limma-moderated t-statistic

```{r}
# Format genes for Reactome pathway analysis, converting to Entrez IDs for fgsea, while handling missing and duplicate Entrez IDs.
ranked_genes <- top_table %>%
  rownames_to_column(var = "ENSEMBL") %>% # Convert rownames to a column.
  arrange(adj.P.Val) %>% # Ensure that the top_table is sorted by BH-adjusted p-value.
  left_join(gene_tool, by = c("ENSEMBL" = "ENSEMBL")) %>% # Add information for Entrez ID and gene symbol.
  dplyr::filter(!is.na(ENTREZID)) %>% # Remove genes without Entrez gene IDs.
  distinct(ENTREZID, .keep_all = TRUE) %>% # In case of duplicates, keep only the first occurrence of each Entrez ID (i.e. the one with the lowest BH-adjusted p-value).
  dplyr::select(ENTREZID, t) %>% # Keep only Entrez IDs and t-statistic.
  arrange(t) %>% # Arrange by t-statistic for GSEA.
  deframe() # Convert to named vector.
```

### Carry out gene set enrichment analysis

```{r}
#| warning: false
# Suppress warning about ties in some gene ranks for GSEA.

# Find Reactome pathways that these genes are present in.
all_pathways <- ranked_genes %>% names(.) %>% reactomePathways(.)
# 1:many mapping reflects that many genes map to multiple Reactome pathways.

# Carry out GSEA analysis, excluding uninformative pathways with over 500 genes.
gsea_results <- fgsea(pathways = all_pathways, stats = ranked_genes, maxSize = 500) %>% 
  arrange(padj)

# Identify pathways that are significantly enriched at a given BH-adjusted α.
enriched_pathways <- gsea_results %>% dplyr::filter(padj < 0.05) %>% 
  arrange(padj)

# Identify shared parent pathways among significantly enriched pathways.
parent_pathway_mapping <- collapsePathways(enriched_pathways, all_pathways, ranked_genes)
parent_enriched_pathways <- enriched_pathways %>%
  dplyr::filter(pathway %in% parent_pathway_mapping$mainPathways) %>%
  arrange(padj)

# Tidy up the environment.
rm(list = c("ranked_genes", "all_pathways", "gsea_results", "parent_pathway_mapping"))
```

### Save parent enriched pathways

```{r}
# Save parent_enriched_pathways as .rds.
saveRDS(parent_enriched_pathways, file = "2_output/parent_enriched_pathways.rds")

# Create a copy of parent_enriched_pathways for publication.
parent_enriched_pathways_pub <- parent_enriched_pathways

# Keep specified columns and reorder.
parent_enriched_pathways_pub <- parent_enriched_pathways_pub %>% 
  dplyr::select(pathway, size, padj, leadingEdge)

# Keep only pathways that are significant at BH-adjusted α < 0.05.
parent_enriched_pathways_pub <- parent_enriched_pathways_pub[parent_enriched_pathways_pub$padj < 0.05, ]

# Convert genes to character.
parent_enriched_pathways_pub$leadingEdge <- as.character(parent_enriched_pathways_pub$leadingEdge)
for (i in 1:nrow(parent_enriched_pathways_pub)) {
  # Split each element in the leadingEdge column.
  genes <- unlist(strsplit(parent_enriched_pathways_pub$leadingEdge[i], "::"))
  # Convert from Entrez IDs to gene symbols.
  genes <- as.numeric(unlist(strsplit(gsub("c\\(|\\)|\"|\\s", "", parent_enriched_pathways_pub$leadingEdge[i]), ",")))
  genes <- gene_tool$SYMBOL[match(genes, gene_tool$ENTREZID)]
  # Optionally sort leading edge alphabetically.
  # genes <- sort(genes)
  # Rejoin.
  parent_enriched_pathways_pub$leadingEdge[i] <- paste(genes, collapse = ", ")
}

# Round adjusted p-value to scientific notation with 3 decimal places.
parent_enriched_pathways_pub$padj <- formatC(parent_enriched_pathways_pub$padj, format = "e", digits = 3)

# Rename columns
parent_enriched_pathways_pub <- dplyr::rename(parent_enriched_pathways_pub, "Reactome Pathway" = "pathway",  "No. of Genes in Pathway" = "size", "Adjusted P-Value" = "padj", "Leading Edge Genes" = "leadingEdge")

# Save table for supplementary materials (including full list of leading edge genes).
parent_enriched_pathways_sup <- parent_enriched_pathways_pub
write.csv(parent_enriched_pathways_sup, file = "2_output/parent_enriched_pathways_sup.csv")

# Remove leading edge genes and save table for in-text.
parent_enriched_pathways_pub <- dplyr::select(parent_enriched_pathways_pub, -"Leading Edge Genes")
write.csv(parent_enriched_pathways_pub, file = "2_output/parent_enriched_pathways_pub.csv")

# Tidy up the environment.
rm(list = c("i", "genes"))
```

### Tidy up the environment and R memory

```{r}
rm(list = ls())
gc()
```
