---
title: "Gene Set Enrichment Analysis"
author: "Kieran Redpath"
format: html
editor: source
theme: cosmo
---


## Summary
+ Reloads processed data from 0_Setup.qmd and 1_Differential_Expression_Analysis.qmd
+ Carries out gene set enrichment analysis for Reactome pathways on differentially expressed gene list, ranked by limma-moderated t-statistic.
+ Collapses Reactome pathway terms to shared parent pathway terms.
+ Saves these outputs as .rds files for later use.

### Load packages
```{r}
#| results: hide
#| warning: false
#| message: false
#| echo: false

library(CePa)
library(data.table)
library(dplyr)
library(edgeR)
library(fgsea)
library(ggplot2)
library(limma)
library(reactome.db)
set.seed(42)
```

### Load pre-processed data
```{r}
# Tool for converting gene names without calling AnnotationDbi::select() every time.
gene_tool <- readRDS("data/gene_tool.rds")

# Tool for indexing cell lines to tissues.
tissue_tool <- readRDS("data/tissue_tool.rds")

# Top table from differential expression analysis.
top_table <- readRDS("1_output/top_table.rds")
```

### Rank genes by limma-moderated t-statistic
```{r}
# Format genes for Reactome pathway analysis, converting to Entrez IDs for fgsea, while handling missing and duplicate Entrez IDs.
# ranked_genes <- top_table %>%
#   rownames_to_column(var = "ENSEMBL") %>% # Convert rownames to a column.
#   arrange(adj.P.Val) %>% # Ensure that the top_table is sorted by BH-adjusted p-value.
#   left_join(gene_tool, by = c("ENSEMBL" = "ENSEMBL")) %>% # Add information for Entrez ID and gene symbol.
#   dplyr::filter(!is.na(ENTREZID)) %>% # Remove genes without Entrez gene IDs.
#   distinct(ENTREZID, .keep_all = TRUE) %>% # In case of duplicates, keep only the first occurrence of each Entrez ID (i.e. the one with the lowest BH-adjusted p-value).
#   dplyr::select(ENTREZID, t) %>% # Keep only Entrez IDs and t-statistic.
#   arrange(t) %>% # Arrange by t-statistic for gsea.
#   deframe() # Convert to named vector.

# Format genes for Reactome pathway analysis, converting from Ensembl to Entrez ID's, and removing any that don't have Entrez ID's.
ranked_genes <- top_table %>%
  rownames_to_column(var = "rn") %>%
  arrange(t) %>%
  left_join(gene_tool, by = c("rn" = "ENSEMBL")) %>%
  dplyr::filter(!is.na(ENTREZID)) %>% # Remove genes without Entrez gene IDs for fgsea.
  dplyr::select(ENTREZID, t) %>%
  deframe()
```

### Carry out gene set enrichment analysis
```{r}
#| warning: false
# Suppress warning about ties in some gene ranks for GSEA analysis.

# Find Reactome pathways that these genes are present in.
all_pathways <- ranked_genes %>% names(.) %>% reactomePathways(.)
# 1:many mapping reflects that many genes map to multiple reactome terms.

# Carry out GSEA analysis, excluding uninformative pathways with over 500 genes.
gsea_results <- fgsea(pathways = all_pathways, stats = ranked_genes, maxSize = 500) %>% 
  arrange(padj) # Dataset contains some duplicate Entrez gene IDs.

# Identify pathways that are significantly enriched at a given BH-adjusted α.
enriched_pathways <- gsea_results %>% dplyr::filter(padj < 0.05) %>% 
  arrange(padj)

# Identify shared parent pathways among significantly enriched pathways.
parent_pathway_mapping <- collapsePathways(enriched_pathways, all_pathways, ranked_genes)
parent_enriched_pathways <- enriched_pathways %>%
  dplyr::filter(pathway %in% parent_pathway_mapping$mainPathways) %>%
  arrange(padj)

# Tidy up the environment.
rm(list = c("ranked_genes", "all_pathways", "gsea_results", "parent_pathway_mapping"))
```

### Save processed objects as .rds files
```{r}
# enriched_pathways
saveRDS(enriched_pathways, file = "2_output/enriched_pathways.rds")

# parent_enriched_pathways
saveRDS(parent_enriched_pathways, file = "2_output/parent_enriched_pathways.rds")
```

### Tidy up the environment and R memory
```{r}
rm(list = ls())
gc()
```
